---
title: "Spatial Ecology demo"
author: "Vasilis Dakos"
date: "2 November 2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE, echo=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
library(knitr)
opts_knit$set(cache=TRUE, autodep=TRUE)
library(purrr)
library(dplyr)
library(broom)
library(ggplot2)
library(rootSolve)
library(numDeriv)
library(tidyr)
library(deSolve)
library(graphics)
library(fields)
```


## The Levins metapopulation model
1. Write the function of the Levins metapopulation model.

```{r echo=TRUE}

levins <- function(t, y, parms) {
  p <- y[1]
  with(as.list(parms), {
    dp <- c * p * (1 - p) - e * p
    return(list(dp))
  })
}
```

By using **with**, we specify the parameters by their names, as long as **parms** includes names. We then use **levins** in the numerical integration function **ode** in the **deSolve** package to simulate its dynamics.

```{r echo=TRUE}
prms <- c(c = 0.15, e = 0.05)
Initial.p <- 0.01
out.L <- data.frame(ode(y = Initial.p, times = 1:100, func = levins, parms = prms))
plot(out.L[, 2] ~ out.L[, 1], type = "l", ylim = c(0, 1), ylab = "p", xlab = "time")
```

2. Find equilibria and for the internal one estimate the bifurcation diagram for changes in colonisation and extinction. Combine the two processes as ratio $c/e = m$.
```{r echo=TRUE}
peq <- expression(ifelse((m - 1)/m<0,0,(m - 1)/m))
m <- seq(0,2,length.out = 20)
plot(m, eval(peq), type = "b", col =1, ylab = "p*", xlab = "m (c/e)")
```

#Mainland-island model (also propagule rain)
$\frac{dp}{dt} = ce(1-p) - ep$

3. Write the full model assuming local colonisation and colonisation from mainland ($c_m$)
4. Plot the colonisation and extinction rates assuming the local colonisation is much smaller than colonisation from mainland. Which model are we recovering? 
5. Find equilibria.
6. Construct the bifurcation diagram or changes in colonisation and extinction (assume  $c/e = m$).
```{r echo=TRUE}
peq <- expression(ifelse(m/(m+1)<0,0,m/(m+1)))
m <- seq(0,2,length.out = 20)
plot(m, eval(peq), type = "b", col =1, ylab = "p*", xlab = "c/e")
```

##Habitat loss scenario
$\frac{dp}{dt} = ce(1-D-p) - ep$

7. Let s assume that habitat is destroyed. What of the two processes will be affected?
How can we incorporate this in the model?
8. Find equilibrium
$P^* = 1 - D - \frac{e}{c}$

9. What is the effect of the amount of habitat destruction on equilibrium?
Show the combined effect of habitat destruction and $c/e = m$ ratio.
Plot $P^*$ as function of $D$ for different values of $m$.
```{r echo=TRUE}
peq <- expression(ifelse(1 - D - 1/m<0,0,1 - D - 1/m))
# m can be 1<m , m<1 always extinction
D <- seq(0,1,length.out = 20)
plot(D, with(as.list(c(m=10,D)),{eval(peq)}), type = "b", col =1, ylab = "p*", xlab = "habitat destroyed, D")
lines(D, with(as.list(c(m=2,D)),{eval(peq)}), type = "b", col =2)
lines(D, with(as.list(c(m=1,D)),{eval(peq)}), type = "b", col =3)
legend("topright", c("m=10", "m=2","m=1"), lty= 1,col = c(1,2,3), bty = "n")
```


#Rescue effect (core-satellite model)
$\frac{dp}{dt} = cp(1-p) - ep(1-ap)$

10. Which process would the size of the patches affect? 
11. But in the Levins model all patches are the same? How can we overcome that to incorporate this in the model?
12. Such effect is termed rescue effect, the strenght of which is $a$. Find the equilibria of the rescue effect model.
The equilibrium is: 
$p^*=\frac{c-e}{c-ae}$
Replace $m=c/e$

13. How is the rescue effect $a$ affecting the equilibrium $P^*$?
```{r echo=TRUE}
peq <- expression(ifelse((m - 1)/(m - a)<0,0,(m - 1)/(m - a)))
m <- seq(0.9,2,length.out = 20)
plot(m, with(as.list(c(m,a=1)),{eval(peq)}), type = "b", col =1, ylab = "p*", xlab = "m",ylim = c(0,1))
lines(m, with(as.list(c(m,a=0.6)),{eval(peq)}), type = "b", col =2)
lines(m, with(as.list(c(m,a=0)),{eval(peq)}), type = "b", col =3)
legend("topleft", c("a=1", "a=0.6","a=0"), lty= 1,col = c(1,2,3), bty = "n")
```

##Rescue effect with habitat loss scenario
$\frac{dp}{dt} = ce(1-D-p) - ep(1-ap)$

14. Retrieve equilibrium 
The equilibrium is given by:
$\frac{(1-D)+\frac{e}{c}}{1-a\frac{e}{c}}$

15. Rewrite in terms of $m=c/e$.
Explore the relative effect of rescue effect and habitat destruction for $m>1$ and $m<1$
What do you observe?

```{r echo=TRUE}
peq <- expression(ifelse((1-m*(1-D))/(a-m)<0,0,(1-m*(1-D))/(a-m)))
D <- seq(0,1,length.out = 40)
m=3
plot(D, with(as.list(c(m,a=0.1)),{eval(peq)}), type = "b", col =1, ylab = "p*", xlab = "habitat destroyed, D",ylim = c(0,1))
lines(D, with(as.list(c(m,a=0.5)),{eval(peq)}), type = "b", col =2)
lines(D, with(as.list(c(m,a=0.9)),{eval(peq)}), type = "b", col =3)
legend("topleft", c("a=0.1", "a=0.5","a=0.9"), lty= 1,col = c(1,2,3), bty = "n")
title("m=3")

m=1.5
plot(D, with(as.list(c(m,a=0.1)),{eval(peq)}), type = "b", col =1, ylab = "p*", xlab = "habitat destroyed, D",ylim = c(0,1))
lines(D, with(as.list(c(m,a=0.5)),{eval(peq)}), type = "b", col =2)
lines(D, with(as.list(c(m,a=0.9)),{eval(peq)}), type = "b", col =3)
legend("topleft", c("a=0.1", "a=0.5","a=0.9"), lty= 1,col = c(1,2,3), bty = "n")
title("m=1.5")
```


## Rescue effect and extinction debt 
16. The above rescue effect seems too small.
Let see it from a different perspective.

Write the Rescue effect model 
```{r echo=TRUE}
rescue_effect <- function(t, y, parms) {
  p <- y[1]
  with(as.list(parms), {
    dp <- c * p * (1 - p) - e * p * (1 - a*p)
    return(list(dp))
  })
}
```

17. Simulate the Levins model with rescue effect and without.
Try these different scenarios:
a. $c<e, a=0.9$
b. $c<e, a=0$

What is the final outcome?
Is the route to extinction the same?
Explain what you think is the “extinction debt”?

```{r echo=TRUE}
t <- 1:200
C1 <- ode(y = 0.999, times = t, func = rescue_effect, parms = c(c = 0.2,  e = 0.25, a = 1))
C2 <- ode(y = 0.99, times = t, func = levins, parms = c(c = 0.2,  e = 0.25, a = 1))

matplot(t, cbind(C1[, 2], C2[, 2]), type = "l", ylab = "p", xlab = "Time", col = 1)
legend("right", c("c < e (rescue)", "c < e (Levins)"), lty = 1:3,  bty = "n")
```

##Cellular automata - an approximation of the Levins model
18. Build a CA model to simulate the Levins model.
Use parameters as in the implicit version.
Create a landscape of 100x100 patches.
Initialise randomly 100 patches.

```{r echo=TRUE}
CALevins <- function(N,e,c,n) {
iN <- numeric()
for (t in 1:n){
  # First estimate survival/extinction
  S <-matrix(runif(10000),nrow=100) 
  N<-N*(S<s)
  
  # Second estimate migrants per patch
  im<-floor(sum(N*c))
  randomly_seeded<-matrix(sample(c(rep(1,im) ,rep (0,10000-im))),nrow=100) 
  N<-N+randomly_seeded
  N<-apply(N,2,function(x) ifelse(x>1,1,x))
  iN[t] <- sum(N)/dim(N)[1]^2
}  
return(list(N=N,iN=iN))
}


#Initialise
c<-0.15 
e<-0.05
s<-(1-e) # survival probability 

#Seed matrix
alpha<-0.2
N <- matrix(sample.int(2,size = 100*100, replace = TRUE,prob=c(alpha,1-alpha)),100,100) 
N[N==2]<-0

#Simulation
n <- 200
ca_Levins <- CALevins(N,e,c,n)
image(1:100,1:100,ca_Levins$N, col=c(0,1)) 
box(col="black")
plot((1:n),ca_Levins$iN,type="l",col=1,xlab = "time")

```

19. Compare the simulated CA equilibrium to the analytical one for a range of c.
What do you find?

```{r echo=TRUE}
c=seq(0.05,0.9,length.out = 10)
e<-0.05
s<-(1-e) # survival probability 

# equilibrium analytical
peq_Levin <- expression((c-e)/(c))
analytical_Peq <- with(as.list(c),{eval(peq_Levin)})

# equilibrium CA
#Initialise
alpha<-0.2
N <- matrix(sample.int(2,size = 100*100, replace = TRUE,prob=c(alpha,1-alpha)),100,100) 
N[N==2]<-0
numerical_Peq<-numeric()
i<-1
n<-1000
ci = seq(from = 0.05,to = 0.9,length.out = 10)
for (c in ci) {
  X <- CALevins(N,e,c,n)
  numerical_Peq[i] <- X$iN[length(X$iN)]
  i<-i+1
}
plot(ci,analytical_Peq,col=1,type = "b",ylim = c(0,1))
lines(ci,numerical_Peq,type = "b",col=2)
legend("right",c("analytical Levins","numerical CA"), col = c(1:2),  bty = "n")
```

##Synchronisation
20. Assume a metapopulation of 2 patches where each population follows Ricker dynamics.
Recall that the Ricker model is:
$N_{t+1} = N_t\exp{(r(1-\frac{N_t}{K})}$

21. Assume local noise and the same carrying capacity $K=1$.
Start with some initial population abundance in both patches below carrying capacity $N_1=0.9, N_2=0.8$  Use an $r=1$.

Also define dispersal $d$ as the fraction of the population that moves between the 2 patches:
```{r echo=TRUE}

ricker <- function(t, y, parms) {
  N0 <- y
  with(as.list(parms), {
    r=matrix(c(r1,r2),nrow = 2,ncol = 1)
    noise = matrix(rnorm(2,sd=sigma),nrow=2,ncol=1)
    N1 <- N0 * exp(r * (1 - N0/K) + noise) # growth
    N1 <- N1 + d*(matrix(rev(N1)-N1,2))/2 # dispersal
    N0 <- N1
    return(list(N1))
  })
}
```

22. Check the simulation for the 2 patches, with $d=0$:
```{r echo=TRUE}
prms <- c(r1 = 1, r2 = 1, K = 3, sigma = 0.01, d = 0) 
Initial.N <- matrix(c(0.9,0.8),2,1)
out.R <- data.frame(ode(y = Initial.N, times = 1:200, func = ricker, parms = prms, method = "iteration"))
colnames(out.R)<-c("time","N1","N2")
ggplot(data = out.R, aes(time)) +
  geom_line(aes(y = N1, color = "N1"))+
  geom_line(aes(y = N2, color = "N2"))
```

23. Now lets add a dispersal rate of $d = 0.5$.
```{r echo=TRUE}
prms <- c(r1 = 1, r2 = 1, K = 3, sigma = 0.01, d = 0.8) 
out.R <- data.frame(ode(y = Initial.N, times = 1:200, func = ricker, parms = prms, method = "iteration"))
colnames(out.R)<-c("time","N1","N2")
ggplot(data = out.R, aes(time)) +
  geom_line(aes(y = N1, color = "N1"))+
  geom_line(aes(y = N2, color = "N2"))
```

24. Lets assume periodic dynamics $r=2$.
compare simulations with $d = 0$ and $d = 0.15$ (what do you observe?)
```{r echo=TRUE}
ro = 2
prms <- c(r1 = ro, r2 = ro, K = 3, sigma = 0.01, d = 0) 
out.R <- data.frame(ode(y = Initial.N, times = 1:100, func = ricker, parms = prms, method = "iteration"))
colnames(out.R)<-c("time","N1","N2")
ggplot(data = out.R, aes(time)) +
  geom_line(aes(y = N1, color = "N1"))+
  geom_line(aes(y = N2, color = "N2"))
prms <- c(r1 = ro, r2 = ro, K = 3, sigma = 0.01, d = 0.5) 
out.R <- data.frame(ode(y = Initial.N, times = 1:100, func = ricker, parms = prms, method = "iteration"))
colnames(out.R)<-c("time","N1","N2")
ggplot(data = out.R, aes(time)) +
  geom_line(aes(y = N1, color = "N1"))+
  geom_line(aes(y = N2, color = "N2"))
```

25. Lets assume chaotic dynamics $r=3$ without noise ($sigma = 0$).
compare simulations with $d = 0$ and $d = 0.3$ (what do you observe?)
```{r echo=TRUE}
ro = 3
Initial.N <- matrix(c(1,2),2,1)
prms <- c(r1 = ro, r2 = ro, K = 3, sigma = 0.00, d = 0) 
out.R <- data.frame(ode(y = Initial.N, times = 1:100, func = ricker, parms = prms, method = "iteration"))
colnames(out.R)<-c("time","N1","N2")
ggplot(data = out.R, aes(time)) +
  geom_line(aes(y = N1, color = "N1"))+
  geom_line(aes(y = N2, color = "N2"))
prms <- c(r1 = ro, r2 = ro, K = 3, sigma = 0.00, d = 0.3) 
out.R <- data.frame(ode(y = Initial.N, times = 1:100, func = ricker, parms = prms, method = "iteration"))
colnames(out.R)<-c("time","N1","N2")
ggplot(data = out.R, aes(time)) +
  geom_line(aes(y = N1, color = "N1"))+
  geom_line(aes(y = N2, color = "N2"))+
  geom_line(aes(y = N2+N1, color = "Total N"))
```
26. Examine the effect of dispersal strength across different values of $r$ (recall $r<2$ is stable, $r>2.56$ chaos).
Measure the correlation between the produced time series as measure of synchrony.
```{r echo=TRUE}
Initial.N <- matrix(c(0.9,0.8),2,1)
ro = seq(1,3,length.out = 20)
n<-500
i<- 1
syn<-matrix(rep(0,20*2),ncol=2)
for (r in ro){
  prms <- c(r1 = r, r2 = r, K = 3, sigma = 0.0001, d = 0)
  out.R <- data.frame(ode(y = Initial.N, times = 1:n, func = ricker, parms = prms, method = "iteration"))
colnames(out.R)<-c("time","N1","N2")
x <- cor.test(out.R$N1[(n/2):n],out.R$N2[(n/2):n],method = "kendall")
syn[i,1]<-x$estimate
i <- i+1
}
i<- 1
for (r in ro){
  prms <- c(r1 = r, r2 = r, K = 3, sigma = 0.0001, d = 0.3)
  out.R <- data.frame(ode(y = Initial.N, times = 1:n, func = ricker, parms = prms, method = "iteration"))
colnames(out.R)<-c("time","N1","N2")
x <- cor.test(out.R$N1[(n/2):n],out.R$N2[(n/2):n],method = "kendall")
syn[i,2]<-x$estimate
i <- i+1
}
matplot(ro, cbind(syn),type = "b",col=c(1,2),xlab = "growth rate r", ylab = "correlation (synchrony)")
legend("topleft", c("d=0", "d=0.3"), lty = 1:2,  bty = "n")
```