---
title: "Spatial Ecology demo to follow"
author: "Vasilis Dakos"
date: "2 November 2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE, echo=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
library(knitr)
opts_knit$set(cache=TRUE, autodep=TRUE)
library(purrr)
library(dplyr)
library(broom)
library(ggplot2)
library(rootSolve)
library(numDeriv)
library(tidyr)
library(deSolve)
library(graphics)
library(fields)
```


## The Levins metapopulation model
1. Write the function of the Levins metapopulation model.
$\frac{dp}{dt} = cp(1-p) - ep$

By using **with**, we specify the parameters by their names, as long as **parms** includes names. We then use **levins** in the numerical integration function **ode** in the **deSolve** package to simulate its dynamics.

```{r echo=FALSE, results="hide"}

```

2. Find equilibria and for the internal one estimate the bifurcation diagram for changes in colonisation and extinction. Combine the two processes as ratio $c/e = m$.
```{r echo=FALSE, results="hide"}

```

##Mainland-island model (also propagule rain)
$\frac{dp}{dt} = ce(1-p) - ep$

3. Write the full model assuming local colonisation and colonisation from mainland ($c_m$)
4. Plot the colonisation and extinction rates assuming the local colonisation is much smaller than colonisation from mainland. Which model are we recovering? 
5. Find equilibria.
6. Construct the bifurcation diagram or changes in colonisation and extinction (assume  $c/e = m$).
```{r echo=FALSE, results="hide"}

```

##Habitat loss scenario
<!-- $\frac{dp}{dt} = ce(1-D-p) - ep$ --> 

7. Let s assume that habitat is destroyed. What of the two processes will be affected?
How can we incorporate this in the model?
8. Find equilibrium
9. What is the effect of the amount of habitat destruction on equilibrium?
Show the combined effect of habitat destruction and $c/e = m$ ratio.
Plot $P^*$ as function of $D$ for different values of $m$.
```{r echo=FALSE, results="hide"}

```


##Rescue effect (core-satellite model)
<!-- $\frac{dp}{dt} = cp(1-p) - ep(1-ap)$ -->

10. Which process would the size of the patches affect? 
11. But in the Levins model all patches are the same? How can we overcome that to incorporate this in the model?
12. Such effect is termed rescue effect, the strenght of which is $a$. Find the equilibria of the rescue effect model. Replace $m=c/e$

13. How is the rescue effect $a$ affecting the equilibrium $P^*$?
```{r echo=FALSE, results="hide"}

```

##Rescue effect with habitat loss scenario
<!-- $\frac{dp}{dt} = ce(1-D-p) - ep(1-ap)$ -->

14. Retrieve equilibrium 
15. Rewrite in terms of $m=c/e$.
Explore the relative effect of rescue effect and habitat destruction for $m>a$ and $m<a$
What do you observe?
```{r echo=FALSE, results="hide"}

```


##A host-parasite metapopulation model
16. Write a host=parasite metapopulation model.
Assume that: a) all patches equally connected.
b) all conspecifics equal risk of extinction
c) colonization proportional to occupied empty pathces
d) host-parasite pathces do not contribute to colonization of empty patches
e) parasite can only occupy patches where a host is present
Find equilibria.
What could be a way to reduce the presence of the parasite?

##Cellular automata - an approximation of the Levins model
17. Study the CA model to simulate the Levins model.
Can you explain each step?
Use parameters as in the implicit version.
Create a landscape of 100x100 patches.
Initialise randomly 100 patches.

```{r echo=TRUE}
CALevins <- function(N,e,c,n) {
iN <- numeric()
for (t in 1:n){
  # First estimate survival/extinction
  S <-matrix(runif(10000),nrow=100) 
  N<-N*(S<s)
  
  # Second estimate migrants per patch
  im<-floor(sum(N*c))
  randomly_seeded<-matrix(sample(c(rep(1,im) ,rep (0,10000-im))),nrow=100) 
  N<-N+randomly_seeded
  N<-apply(N,2,function(x) ifelse(x>1,1,x))
  iN[t] <- sum(N)/dim(N)[1]^2
}  
return(list(N=N,iN=iN))
}


#Initialise
c<-0.15 
e<-0.05
s<-(1-e) # survival probability 

#Seed matrix
alpha<-0.2
N <- matrix(sample.int(2,size = 100*100, replace = TRUE,prob=c(alpha,1-alpha)),100,100) 
N[N==2]<-0

#Simulation
n <- 200
ca_Levins <- CALevins(N,e,c,n)
image(1:100,1:100,ca_Levins$N, col=c(0,1)) 
box(col="black")
plot((1:n),ca_Levins$iN,type="l",col=1,xlab = "time")

```

18. Compare the simulated CA equilibrium to the analytical one for a range of c.
What do you find?
```{r echo=FALSE}

```

##Synchronisation
19. Assume a metapopulation of 2 patches where each population follows Ricker dynamics.
Recall that the Ricker model is:
$N_{t+1} = N_t\exp{(r(1-\frac{N_t}{K})}$

20. Assume local noise and the same carrying capacity $K=1$.
Start with some initial population abundance in both patches below carrying capacity $N_1=0.9, N_2=0.8$  Use an $r=1$.

Also define dispersal $d$ as the fraction of the population that moves between the 2 patches:
```{r echo=FALSE}

```

21. Check the simulation for the 2 patches, with $d=0$:
```{r echo=FALSE, results="hide"}

```

22. Now lets add a dispersal rate of $d = 0.5$.
```{r echo=FALSE, results="hide"}

```

23. Lets assume periodic dynamics $r=2$.
compare simulations with $d = 0$ and $d = 0.15$ (what do you observe?)
```{r echo=FALSE, results="hide"}

```

24. Lets assume chaotic dynamics $r=3$ without noise ($sigma = 0$).
compare simulations with $d = 0$ and $d = 0.3$ (what do you observe?)
```{r echo=FALSE, results="hide"}

```
25. Examine the effect of dispersal strength across different values of $r$ (recall $r<2$ is stable, $r>2.56$ chaos).
Measure the correlation between the produced time series as measure of synchrony.
```{r echo=FALSE, results="hide"}

```